---
title: "golem"
author: "Stephen J. Price"
date: "15/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Resources
- [workshop flyer](https://f.hubspotusercontent00.net/hubfs/3463003/Good%20practices%20for%20%7Bshiny%7D%20development%20using%20%7Bgolem%7D%20-%20An%20Introduction.pdf)
- [github repo](https://github.com/ColinFay/earlworkshop2020)
- [slides](https://speakerdeck.com/colinfay/workshop-good-practices-for-shiny-development-with-golem)
- [golem book](https://engineering-shiny.org/)
- [basic exercises](https://connect.thinkr.fr/make-a-golem/) - copied below
- [shiny developer series: golem](https://shinydevseries.com/post/episode-2-golem/)

# Make a {golem}
## Intro
### Ex 1: Install the latest version
```{r}
remotes::install_github("Thinkr-open/golem")
```

### Ex 2: Create a new {golem} project
File > New project > New directory > Package for shiny app using golem.  
Note: it’s possible that you will need to restart RStudio after the installation.  

## 01_start.R
you need to set and run everything in this doc at the start of the project but never again `r emo::ji("smile")`. Modify it as needed and run.  

### Ex 3: Fill the DESCRIPTION using fill_dec()

### Ex 4.1: Add a README, the GPL3 licence, and a NEWS.md

### Ex 4.2: Use the recommended tests, recommended deps, utils_ui and utils_server
Try to launch your app with `golem::run_dev()`.  

## 02_dev.R
### Ex 5: Add a dataset
Add a dataset, by running `usethis::use_data_raw( name = "my_dataset" )` from line 38 of 02_dev.R.  

Inside the script that opens, write:
```{r eval=FALSE}
my_dataset <- readr::read_csv("https://raw.githubusercontent.com/ColinFay/earlworkshop2020/main/data/dataset.csv")
usethis::use_data(my_dataset, overwrite = TRUE)

```

The dataset is now added to your app

### Ex 6: Building your first module
- Add two dependencies to your app, {dplyr} and {ggplot2}, by using the `usethis::use_package` function. These packages will be used inside our module.
- Add the pipe to your app with `usethis::use_pipe()`
- Create a module with `golem::add_module`
  + the script that's created even contains the code (commented at the bottom) that you in your app_ui and app_server
  + Just copy and paste these UI and Server functions to app_ui and app_server
- Reproduce this UI/Server inside your module
- Try to launch your app with `golem::run_dev()`

### Ex 7 :CSS
Create a new CSS file, and change the background of the app to LightGreen

### Ex 8: Deploy
Run one of:
- `golem::add_rstudioconnect_file()`
- `golem::add_shinyappsio_file()`
- `golem::add_shinyserver_file()`

Depending on where you want to deploy your app.
